---
title: "README"
author: "Michelle Miron"
date: "10/10/2017"
output:
  pdf_document: default
  html_document: default
---



```{r setup, include=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.height = 10, fig.width = 14)
library(dplyr)
library(tidyr)
library(ggplot2)
library(gapminder)
library(magrittr)
library(gplots)

```



###Overview

### Load data
This is what the Raw data looks like:
           
```{r}
          Raw_BM_TEM <- read.csv("data/BM_TEM_D332_333_335_exp2.csv")
          colnames(Raw_BM_TEM)
```

The problem is there is a separate file for each Tissue site (a variable) :

```{r}
          Raw_BM_TEM <- read.csv("data/BM_TEM_D332_333_335_exp2.csv")
          Raw_Spl_TEM <- read.csv("data/SPL_TEM_D332_333_335_exp2.csv")
          Raw_LN_TEM <- read.csv("data/LLN_TEM_D332_333_335_exp2.csv") 
          Raw_Lng_TEM <- read.csv("data/Lung_TEM_D332_333_335_exp2.csv")
          Raw_LN_Naive <- read.csv("data/LLN_Naive_D332_333_335_exp2.csv")
```


In order to fix this, the next step is to merge the data from each file, which cnotained data on all the cells within on subset (Naive or Memory T cells from a particular tissue), into a database/file with all the data from each cell subset from different tissues together. I call the merged data "tidy" data below. 

### Tidy data and Messy data
Tidy [data](http://vita.had.co.nz/papers/tidy-data.html) is a uniform way to make data more organized and easier to do statsitics and analysis on. Here I will put the data into the tidy format. This consists of three steps:

1.Only Keep relevant columns in data set:

```{r}

# A function to keep a subset of columns in the dataframe
# and to fix an aritifact in the File.Identifier column
tidyDataColumns <- function(RawData) {
          
          # Name columns to keep
          keeps <- c("Time","CD57","CD28","CD45RA","CD103","Perforin",
                                         "CD127","PD_1","ICOS","CD27","CCR5","Bcl6",
                                         "Tcf1","CD14","CD56","CXCR5","CD69","X41BB","CD25",
                                         "CCR7","CD3","Tbet","CD38","CD95","LAG3",
                                         "CXCR4","HLADR", "TIGIT","GranzymeB",
                                         "File.Identifier")
        
          # The subset of the raw data that contains columns defined above
          KeyMarkersdf <- RawData[keeps]
                    
          # There is an artifact of the data
          # The "File.Identifier" column was generated from the FCSexpress program
          # This column is a number 1-12 that is a unique identifier 
          # The unique identifier tells you what sample the cells are from
          # For example, 1 = Bome marrrow from D332, 2= BM from D333
          # The key to this identifier is in step 2 of tidy data below
          # The artifact is that the he program outputs the File.Identifier number 
          # as a real number ie. (1.04243) instead of an integer ie. (1)
          # Therefore I rounded the File.Identifier column here:
          File.ID.Rounded <- round(KeyMarkersdf$File.Identifier, digits=0)
          # Take away the non-rounded file identifier
          NoFileIDdf <- subset(KeyMarkersdf, select = -c(File.Identifier) )
          # Add_Rounded_FileID to the dataframe
          NoFileIDdf$File.ID <- File.ID.Rounded
          # Rename the dataset
          Tidydf <- NoFileIDdf
}

# Apply this function to all data sets
          Tidy_BM_TEM <- tidyDataColumns(Raw_BM_TEM)
          Tidy_LN_TEM <- tidyDataColumns(Raw_LN_TEM)
          Tidy_Spl_TEM <- tidyDataColumns(Raw_Spl_TEM)
          Tidy_Lung_TEM <- tidyDataColumns(Raw_Lng_TEM)
          Tidy_LN_Naive <- tidyDataColumns(Raw_LN_Naive)
```

2. Combined the data sets together into one table:

Remember that for this data set, the tissue varibale can still be disintuigshed by the file ID column. However the subset Variable cannot, So i will add this variable to each data set:

Add a column that says subset name as a new variable:

```{r}
Tidy_BM_TEM["Subset"] <- "TEM" 
Tidy_LN_Naive["Subset"] <- "Naive" 
Tidy_LN_TEM["Subset"] <- "TEM" 
Tidy_Lung_TEM["Subset"] <- "TEM" 
Tidy_Spl_TEM["Subset"] <- "TEM" 

```

And now combine each data set by adding new rows:

```{r}
TidyMasterData <- do.call( "rbind", list(Tidy_BM_TEM, 
                                         Tidy_LN_Naive,Tidy_LN_TEM, 
                                         Tidy_Spl_TEM, Tidy_Lung_TEM))
```


3. Add variables that are imbedded in "File.Identifier" column

The "File.Identifier" column has a key which reveals what variables they represent:

```{r}
#add tissue and donor column
TidyMasterData["Tissue"] <- NA
TidyMasterData["Donor"] <- NA
library(dplyr)
                              
#Here is the key for what the File.ID column is
File.Identifier_Key <- read.csv("data/File Identifier key.csv")
File.Identifier_Key

#Use the File.Identifier Key to label the table for the correct variables
#associated with the observations
for (i in 1:length(File.Identifier_Key[,1])) {
          TidyMasterData$Tissue[TidyMasterData$File.ID==i] <- as.character(
                    File.Identifier_Key$Tissue[i])
          TidyMasterData$Donor[TidyMasterData$File.ID==i] <- File.Identifier_Key$Donor[i]      
}

library(dplyr)
sample_n(TidyMasterData, 10)
```

Now to fix order of the columns so that the fixed variables ( File.ID, Subset, Tissue, and Donor) are first, and the oberserved variables (Protein names like CD57, T-bet , etc. are next). Here we will also remove the Time and File.ID column since those are no longer needed.
                    
```{r}
#Remove columns that are not needed by their number in the dataframe
TidyData <- TidyMasterData[,c(33,32,31,2:29)]
sample_n(TidyData, 10)
```

Time to  make the last changes to make tidy data.
Let's add a column to be a unique identifier for each cell number

```{r}             
TidyData["Cell"] <- 1:nrow(TidyData)
TidyDatabyCell<- TidyData[,c(32,1:31)]
messy <- TidyDatabyCell
library(tidyr)
```

This will turn all the marker variables into rows instead of columns

```{r}
tidy <- messy %>%
gather(Marker, Expression, CD57:GranzymeB)
sample_n(tidy,10)
sample_n(messy,10)
                    
```

### Mean of the data 
#used this tutorial: http://stat545.com/block023_dplyr-do.html
```{r}
library(dplyr)
library(gapminder)
#this function allows multiple pipes to use with dplyr
library(magrittr)

meanDB <- tidy %>%
           group_by(Donor,Tissue, Subset, Marker) %>%
                    summarize(meanExpression = mean(Expression))
                    
#and Z score of this data
ZscoreDB <- meanDB %>% 
  group_by(Donor,Marker) %>% 
  mutate(z_score = scale(meanExpression))  
 
ZscoreDB <- as.data.frame(ZscoreDB)
  meanZscoredf <- ZscoreDB
  
```
#create unique name for each sample type
```{r}
MeanZscore_SampleIDdf <- meanZscoredf
MeanZscore_SampleIDdf$Sample <- paste(MeanZscore_SampleIDdf$Donor,MeanZscore_SampleIDdf$Tissue,MeanZscore_SampleIDdf$Subset)
sample_n(MeanZscore_SampleIDdf,10)

```

### plot Z score mean data

Lets first get the data "untidy" so it's easier to use heatmap.2
```{r}
library(tidyr)
Zscoredf <- MeanZscore_SampleIDdf[,c(4,6,7)] 
ZscoreforHeatmap <- spread(Zscoredf, Marker, z_score)
ZscoreforHeatmap_justValues <- ZscoreforHeatmap[,-c(1)]
#name rows of the df for heatmap plotting so labels showup
rownames(ZscoreforHeatmap_justValues) <- ZscoreforHeatmap[,1]

#graph heatmap
#going to try following this [heatmap #tutorial](http://varianceexplained.org/r/love-actually-network/)
#color cheat sheet https://www.nceas.ucsb.edu/~frazier/RSpatialGuides/colorPaletteCheatsheet.pdf
library(gplots)
my_palette <- colorRampPalette(c("steelblue3", "white", "sienna3"))(n = 100)
heatmap.2(as.matrix(t(ZscoreforHeatmap_justValues)),trace="none",col=my_palette,cexRow=0.75, cexCol=0.5, breaks=c(seq(-1,1,0.02)))

#lets remove some rows to only keep markers we want to plot!
colnames(ZscoreforHeatmap_justValues)
SelectZscores <-ZscoreforHeatmap_justValues[,-c(4,6,10,11,13,15,20,21,22,28)]
my_palette <- colorRampPalette(c("steelblue3", "white", "sienna3"))(n = 100)

```


Now lets to some MDS plotting and K means clustering
```{r}

#Using this [tutorial](http://genomicsclass.github.io/book/pages/clustering_and_heatmaps.html) to make some MDS plots and do kMeans

#kmeans on MDS plot
set.seed(1)
d <- dist(ZscoreforHeatmap_justValues)
km <- kmeans(ZscoreforHeatmap_justValues, centers = 2)
names(km)
mds <- cmdscale(d)
plot(mds[,1], mds[,2], col=km$cluster, pch=16 )
text(mds[,1], mds[,2], names(km$cluster),cex=0.5)

#kmeans only on some markers
set.seed(1)
d <- dist(SelectZscores)
km <- kmeans(SelectZscores, centers = 2)
names(km)
mds <- cmdscale(d)
plot(mds[,1], mds[,2], col=km$cluster, pch=16 )
text(mds[,1], mds[,2], names(km$cluster),cex=0.5)

```


